---
output: github_document
---

```{r setup, echo = FALSE, message=FALSE, results='hide'}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.path = "man/figures/README-",
  echo = TRUE,
  fig.width = 8,
  fig.height = 6
)

```

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/incidence2)](https://CRAN.R-project.org/package=incidence2)
[![R-CMD-check](https://github.com/reconverse/incidence2/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/reconverse/incidence2/actions/workflows/R-CMD-check.yaml)
[![](https://raw.githubusercontent.com/reconverse/reconverse.github.io/master/images/badge-stable.svg)](https://www.reconverse.org/lifecycle.html#stable)
<!-- badges: end -->


# incidence2

incidence2 is an R package that implements functions and classes to compute,
handle and visualise *incidence* from linelist data. It refocusses the scope of
the original [incidence](https://cran.r-project.org/package=incidence) package.
Unlike the original package, incidence2 concentrates only on the initial
calculation, manipulation and plotting of the resultant *incidence* objects.

## Installing the package

You can install the released version of incidence2 from [CRAN](https://cran.r-project.org/) with:

```{r, eval=FALSE}
install.packages("incidence2")
```

## Vignettes

An overview of incidence2 is provided in the vignetted distributed with the
package:

* `vignette("incidence2", package = "incidence2")`
* `vignette("working-with-incidence2-objects", package = "incidence2")`
