---
output: github_document
---

```{r setup, echo = FALSE, message=FALSE, results='hide'}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.path = "man/figures/README-",
  echo = TRUE,
  fig.width = 8,
  fig.height = 6
)

```

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/incidence2)](https://CRAN.R-project.org/package=incidence2)
[![Codecov test coverage](https://codecov.io/gh/reconverse/incidence2/branch/master/graph/badge.svg)](https://codecov.io/gh/reconverse/incidence2?branch=master)
[![R-CMD-check](https://github.com/reconverse/incidence2/workflows/R-CMD-check/badge.svg)](https://github.com/reconverse/incidence2/actions)
[![](https://raw.githubusercontent.com/reconverse/reconverse.github.io/master/images/badge-stable.svg)](https://www.reconverse.org/lifecycle.html#stable)
<!-- badges: end -->


# [incidence2](https://www.reconverse.org/incidence2/)
incidence2 is an R package that implements functions and classes to compute, handle and visualise incidences from linelist data.
It refocusses the scope of the original [incidence](https://github.com/reconhub/incidence) package.
Unlike the original package, incidence2 concentrates only on the initial calculation, manipulation and plotting of the resultant incidence objects.

## Installing the package

The development version, which this documentation refers to, can be installed
from [GitHub](https://github.com/) with:

```{r, eval=FALSE}
if (!require(remotes)) 
  install.packages("remotes")
}
remotes::install_github("reconverse/incidence2", build_vignettes = TRUE)
```

You can install the current version of the package from the releases
[page](https://github.com/reconverse/incidence2/releases) or directly from
[CRAN](https://cran.r-project.org/) with:

```{r, eval=FALSE}
install.packages("incidence2")
```

## Resources

### Vignettes

A short overview of incidence2 is provided below in the worked example below.
More detailed tutorials are distributed as vignettes with the package:

* `vignette("Introduction", package = "incidence2")`
* `vignette("handling_incidence_objects", package = "incidence2")`
* `vignette("customizing_incidence_plots", package = "incidence2")`
* `vignette("alternative_date_groupings", package = "incidence2")`

## Getting help online
- Bug reports and feature requests should be posted on *github* using the issue system: <https://github.com/reconverse/incidence2/issues>.
- Online documentation: <https://www.reconverse.org/incidence2>.
- All other questions should be posted on the **RECON** slack channel;
  see <https://www.repidemicsconsortium.org/forum/> for details on how to join.

# A quick overview

This short example uses the simulated Ebola Virus Disease (EVD) outbreak from the package [*outbreaks*](https://github.com/reconhub/outbreaks).
It shows how to compute incidence for various time steps and plot the resulting output.

First, we load the data:

```{r data, message=FALSE}
library(outbreaks)
library(incidence2)

dat <- ebola_sim_clean$linelist
str(dat)
```

## Computing and plotting incidence

We compute the weekly incidence:

```{r incid7}
i_7 <- incidence(dat, date_index = date_of_onset, interval = 7)
i_7
summary(i_7)
plot(i_7, color = "white")
```

Notice how specifying the interval as 7 creates weekly intervals with the coverage displayed by date.
`incidence()` also allows us to create year-weekly groupings with the default being weeks starting on a Monday (following the ISO 8601 date and time standard).
`incidence()` can also compute incidence by specified groups using the `groups` argument.
As an example, below we can compute the weekly incidence by gender and plot in a single, stacked chart:

```{r genderstack}
iw <- incidence(dat, interval = "week", date_index = date_of_onset, groups = gender)
iw
summary(iw)
plot(iw, fill = "gender", color = "white")
```

we can also facet our plot (grouping detected automatically):

```{r genderfacet}
facet_plot(iw, n.breaks = 3, color = "white")
```

It is also possible togroup by multiple variables specifying different facets and fills:

```{r genderhospital}
iw2 <- incidence(dat, date_of_onset, interval = "week",  groups = c(gender, hospital))
iw2
summary(iw2)
facet_plot(iw2, facets = gender, fill = hospital, n.breaks = 3)
```

## Using an alternative function

The `incidence()` function wraps the date grouping functionality of the [grates](https://www.reconverse.org/grates) package, providing an easy to use interface for constructing incidence objects.
Sometimes, however, you may want greater flexibility in choosing how you would like to transform your "date" inputs.
Using the function `build_incidence()`,you can specify the function you wish to apply.
We illustrate this below with the excellent [clock](https://clock.r-lib.org) package:

```{r clock}
library(clock)

# create a week function comparable to above approach
isoweek <- function(x) calendar_narrow(as_iso_year_week_day(x), "week")

clock_week_inci <- 
  build_incidence(
    dat,
    date_index = date_of_onset,
    groups = c(gender, hospital),
    FUN = isoweek
  )

clock_week_inci

