---
output: github_document
---

```{r setup, echo = FALSE, message=FALSE, results='hide'}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.path = "man/figures/README-",
  echo = TRUE,
  fig.width = 8,
  fig.height = 6
)

```

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/incidence2)](https://CRAN.R-project.org/package=incidence2)
[![Codecov test coverage](https://codecov.io/gh/reconhub/incidence2/branch/master/graph/badge.svg)](https://codecov.io/gh/reconhub/incidence2?branch=master)
[![R-CMD-check](https://github.com/reconhub/incidence2/workflows/R-CMD-check/badge.svg)](https://github.com/reconhub/incidence2/actions)
<!-- badges: end -->


# Scope

*incidence2* refocusses the scope of the original
[incidence](https://github.com/reconhub/incidence) package.  The aim is to
provide a "tidy" interface for users to work with whilst at the same time
simplifying the underlying implementation. To this end, *incidence2* 
concentrates only on the initial data handling, calculation and graphing of 
incidence objects. The "fitting" and "peak estimation" functions of 
[incidence](https://github.com/reconhub/incidence) (e.g. `incidence::fit` and
`incidence::estimate_peak`) are being implemented in an accompanying package 
called [`i2extras`](https://github.com/reconhub/i2extras).  Here 
they will have a more consistent interface, better choice of underlying models,
and tidier outputs. 


# Installing the package

The development version, which this documentation refers to, can be installed
from [GitHub](https://github.com/) with:

```{r, eval=FALSE}
if (!require(remotes)) {
  install.packages("remotes")
}
remotes::install_github("reconhub/incidence2", build_vignettes = TRUE)
```

You can install the current version of the package from the releases
[page](https://github.com/reconhub/incidence2/releases) or directly from
[CRAN](https://cran.r-project.org/) with:

```{r, eval=FALSE}
install.packages("incidence2")
```


# Resources

## Vignettes

A short overview of *incidence2* is provided below in the worked example below.
More detailed tutorials are distributed as vignettes with the package:

* `vignette("Introduction", package = "incidence2")`
* `vignette("handling_incidence_objects", package = "incidence2")`
* `vignette("customizing_incidence_plots", package = "incidence2")`

## Websites

The following websites are available:

- The *incidence2* project on *github*, useful for developers, contributors, and
users wanting to post issues, bug reports and feature requests: <br>
<https://github.com/reconhub/incidence2>

## Getting help online

Bug reports and feature requests should be posted on *github* using the
[*issue* system](https://github.com/reconhub/incidence2/issues). All other
questions should be posted on the **RECON** slack channel; see 
<https://www.repidemicsconsortium.org/forum/> for details on how to join.


# A quick overview

This short example uses the simulated Ebola Virus Disease (EVD) outbreak from
the package [*outbreaks*](https://github.com/reconhub/outbreaks). It shows how
to compute incidence for various time steps plot the resulting incidence tables.

First, we load the data:

```{r data, message=FALSE}
library(outbreaks)
library(incidence2)

dat <- ebola_sim_clean$linelist
str(dat)
```

## Computing and plotting incidence

We compute the weekly incidence:

```{r incid7}
i_7 <- incidence(dat, date_index = date_of_onset, interval = 7)
i_7
summary(i_7)
plot(i_7, color = "white")
```

Notice how specifying the interval as 7 creates weekly intervals with the
coverage displayed by date. Below we illustrate how `incidence()` also allows us
to create year-weekly groupings with the default being weeks starting on a
Monday (following the ISO 8601 date and time standard). `incidence()` can also
compute incidence by specified groups using the `groups` argument. For instance,
we can compute the weekly incidence by gender and plot in a single, stacked
chart:

```{r genderstack}
i_week_sex <- incidence(dat, interval = "week", date_index = date_of_onset,
                        groups = gender)
i_week_sex
summary(i_week_sex)
plot(i_week_sex, fill = "gender", color = "white")
```

we can also facet our plot (grouping detected automatically):

```{r genderfacet}
facet_plot(i_week_sex, n_breaks = 3, color = "white")
```

and we can also group by multiple variables specifying different facets and 
fills:

```{r genderhospital}
# incidence is compatible with the magrittr pipe operator
i_week_sh <- incidence(dat, date_index = date_of_onset, interval = "week",
                       groups = c(gender, hospital))
i_week_sh
summary(i_week_sh)
facet_plot(i_week_sh, facets = gender, fill = hospital, n_breaks = 3)
```
