---
output: github_document
---

```{r setup, echo = FALSE, message=FALSE, results='hide'}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  fig.path = "figs/",
  echo = TRUE
)

```

# incidence2

<!-- badges: start -->
[![Project Status: WIP – Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/incidence2)](https://CRAN.R-project.org/package=incidence2)
[![R build status](https://github.com/reconhub/incidence2/workflows/R-CMD-check/badge.svg)](https://github.com/reconhub/incidence2/actions)
<!-- badges: end -->

**Disclaimer:** this package is a work in progress, in constant development. 
Please reach out to the authors before using.

*incidence2* refocusses the scope of the original
[incidence](https://github.com/reconhub/incidence) package.  The aim is to
provide a "tidy" interface for users to work with whilst at the same time
simplifying the underlying implementation.  

Whilst we are in the early stages of development, we aim to re-implement
the "fitting" functions of [incidence](https://github.com/reconhub/incidence)
(e.g. `incidence::fit`) in a smaller, separate, package with a more consistent 
interface, more choice of underlying models, and tidier outputs. To this end,
the package scope of *incidence2* focusses only on the initial data handling,
calculation and graphing of incidence objects.


## Installing the package

To install the development, *github* version of the package use:

```{r install2, eval=FALSE}
devtools::install_github("reconhub/incidence")
```
