---
output: github_document
---

```{r setup, echo = FALSE, message=FALSE, results='hide'}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.path = "man/figures/README-",
  echo = TRUE,
  fig.width = 8,
  fig.height = 6
)

```

<!-- badges: start -->
[![Project Status: WIP – Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/incidence2)](https://CRAN.R-project.org/package=incidence2)
[![R build status](https://github.com/reconhub/incidence2/workflows/R-CMD-check/badge.svg)](https://github.com/reconhub/incidence2/actions)
[![Codecov test coverage](https://codecov.io/gh/reconhub/incidence2/branch/master/graph/badge.svg)](https://codecov.io/gh/reconhub/incidence2?branch=master)
<!-- badges: end -->


# Scope

*incidence2* refocusses the scope of the original
[incidence](https://github.com/reconhub/incidence) package.  The aim is to
provide a "tidy" interface for users to work with whilst at the same time
simplifying the underlying implementation. To this end, *incidence2* 
concentrates only on the initial data handling, calculation and graphing of 
incidence objects. The "fitting" and "peak estimation" functions of 
[incidence](https://github.com/reconhub/incidence) (e.g. `incidence::fit` and
`incidence::estimate_peak`) are being implemented in an accompanying package 
called [`incidence2plus`](https://github.com/reconhub/incidence2extra).  Here 
they will have a more consistent interface, better choice of underlying models,
and tidier outputs. 


# What does it do?

The main features of the package include:

- **`incidence()`**: compute incidence from both linelist and pre-aggregated
  datasets; any fixed time interval can be used; the returned object is a 
  tibble subclass called *incidence2*.

- plotting functions **`plot()`** and **`facet_plot()`**: these functions
  return customised ggplot2 plots of *incidence2* objects 
  (see **`plot.incidence2()`** for details).

- Compatible with [dplyr](https://dplyr.tidyverse.org/) for data manipulation.
  (see `vignette("handling_incidence_objects")` for more details).

- **`regroup()`**: regroup incidence from different groups into one global incidence
  time series.

- **`cumulate()`**: computes cumulative incidence over time from an `incidence()`
  object.

- **`print()`** and **`summary()`** functions.

- Conversion functions:
    
    - **`as.data.frame()`**: converts an `incidence()` object into a `data.frame`.
    
    - **`as_tibble()`**: converts an `incidence()` object into a `tibble`.

- Accessor functions: **`get_counts_name()`**, **`get_dates_name()`**,
  **`get_dates()`**, **`get_group_names()`**, **`get_interval()`**, 
  **`get_timespan()`** and  **`get_n()`**.


# Installing the package

You can install the current version of the package from 
[CRAN](https://cran.r-project.org/) with:

```{r, eval=FALSE}
install.packages("incidence2")
```

The development version can be installed from [GitHub](https://github.com/) with:

```{r, eval=FALSE}
if (!require(remotes)) {
  install.packages("remotes")
}
remotes::install_github("reconhub/incidence2", build_vignettes = TRUE)
```

# Resources

## Vignettes

A short overview of *incidence2* is provided below in the worked example below.
More detailed tutorials are distributed as vignettes with the package:

* `vignette("Introduction", package = "incidence2")`
* `vignette("handling_incidence_objects", package = "incidence2")`
* `vignette("customizing_incidence_plots", package = "incidence2")`

## Websites

The following websites are available:

- The *incidence2* project on *github*, useful for developers, contributors, and
users wanting to post issues, bug reports and feature requests: <br>
<https://github.com/reconhub/incidence2>

## Getting help online

Bug reports and feature requests should be posted on *github* using the [*issue* system](https://github.com/reconhub/incidence2/issues). All other questions
should be posted on the **RECON** slack channel see 
<https://www.repidemicsconsortium.org/forum/> for details on how to join.


# A quick overview

This short example uses the simulated Ebola Virus Disease (EVD) outbreak from
the package [*outbreaks*](https://github.com/reconhub/outbreaks). It shows how
to compute incidence for various time steps plot the resulting incidence tables.

First, we load the data:

```{r data, message=FALSE}
library(outbreaks)
library(incidence2)

dat <- ebola_sim_clean$linelist
str(dat)
```

## Computing and plotting incidence

We compute the weekly incidence:

```{r incid7}
i_7 <- incidence(dat, date_index = date_of_onset, interval = 7)
i_7
summary(i_7)
plot(i_7, color = "black")
```

`incidence()` can also compute incidence by specified groups using the `groups`
argument. For instance, we can compute the weekly incidence by gender and plot
in a single, stacked chart:

```{r genderstack}
i_7_sex <- incidence(dat, interval = "week", 
                     date_index = date_of_onset, groups = gender)
i_7_sex
summary(i_7_sex)
plot(i_7_sex, fill = "gender", color = "black")
```

we can facet our plot (grouping detected automatically):

```{r genderfacet}
facet_plot(i_7_sex, n_breaks = 4)
```

and we can also group by multiple variables specifying different facets and 
fills:

```{r genderhospital}
# incidence is compatible with the magrittr pipe operator
i_7_sh <- 
  dat %>% 
  incidence(date_index = date_of_onset, interval = "week", 
            groups = c(gender, hospital))
i_7_sh
i_7_sh %>% summary()
i_7_sh %>% facet_plot(facets = gender, fill = hospital)
```


