---
output: github_document
---

```{r setup, echo = FALSE, message=FALSE, results='hide'}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  fig.path = "figs/",
  echo = TRUE
)

```

<!-- badges: start -->
[![Project Status: WIP – Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/incidence2)](https://CRAN.R-project.org/package=incidence2)
[![R build status](https://github.com/reconhub/incidence2/workflows/R-CMD-check/badge.svg)](https://github.com/reconhub/incidence2/actions)
[![codecov](https://codecov.io/gh/reconhub/incidence2/branch/master/graph/badge.svg)](https://codecov.io/gh/reconhub/incidence2)
<!-- badges: end -->

**Disclaimer:** this package is a work in progress. Please reach out to the
authors before using.

# Scope

*incidence2* refocusses the scope of the original
[incidence](https://github.com/reconhub/incidence) package.  The aim is to
provide a "tidy" interface for users to work with whilst at the same time
simplifying the underlying implementation. To this end, *incidence2* 
concentrates only on the initial data handling, calculation and graphing of 
incidence objects. The "fitting" functions of 
[incidence](https://github.com/reconhub/incidence) (e.g. `incidence::fit`) 
will likely find there place in a smaller, separate, package with a more 
consistent interface, more choice of underlying models, and tidier outputs. 


# Installing the package

To install the development, *github* version of the package use:

```{r install2, eval=FALSE}
devtools::install_github("reconhub/incidence2")
```


# What does it do?

The main features of the package include:

- **`incidence()`**: compute incidence from linelist datasets; any fixed
  time interval can be used; the returned object is a tibble subclass called
  *incidence*.

- plotting functions **`plot()`** and **`facet_plot()`**: these functions
  return customised ggplot2 plots of *incidence* objects 
  (see `?plot.incidence` for details).

- Compatible with [dplyr](https://dplyr.tidyverse.org/) for data manipulation.

- **`regroup()`**: regroup incidence from different groups into one global incidence
  time series.

- **`cumulate()`**: computes cumulative incidence over time from an `incidence`
  object.
  
- **`bootstrap()`**: generates a bootstrapped *incidence()* object by re-sampling,
  with replacement, the original dates of events.

- **`find_peak()`**: locates the peak time of the epicurve.

- **`estimate_peak()`**: uses bootstrap to estimate the peak time (and related
  confidence interval) of a partially observed outbreak.
  
- **`print()`** and **`summary()`** functions.

- Conversion functions:
    
    - **`as.data.frame()`**: converts an `incidence()` object into a `data.frame`.
    
    - **`as_tibble()`**: convertss an `incidence()` object into a `tibble`.

- Accessor functions: **`get_count_vars()`**, **`get_date_vars()`**,
  **`get_group_vars()`**, **`get_interval()`**, **`get_timespan()`** and
  **`get_n()`**.


# A quick overview

The following worked example provides a brief overview of the package's
functionalities.

## Loading the data
This example uses the simulated Ebola Virus Disease (EVD) outbreak from the
package [*outbreaks*](https://github.com/reconhub/outbreaks). It shows how to
compute incidence for various time steps and manipulate the data.

First, we load the data:

```{r data, message=FALSE}
library(outbreaks)
library(incidence2)
library(dplyr)

dat <- ebola_sim_clean$linelist
glimpse(dat)
```

## Computing and plotting incidence

We compute the weekly incidence:

```{r incid7}
i_7 <- incidence(dat, date_index = date_of_onset, interval = 7)
i_7
summary(i_7)
plot(i_7, color = "black")
```

`incidence()` can also compute incidence by specified groups using the `groups`
argument. For instance, we can compute the weekly incidence by gender and plot
in a single, stacked chart:

```{r genderstack}
i_7_sex <- incidence(dat, interval = "week", 
                     date_index = date_of_onset, groups = gender)
i_7_sex
summary(i_7_sex)
plot(i_7_sex, fill = "gender", color = "black")
```

we can facet our plot (grouping detected automatically):

```{r genderfacet}
facet_plot(i_7_sex, n_breaks = 4)
```

and we can also group by multiple variables specifying different facets and 
fills

```{r genderhospital}
# incidence is compatible with the magrittr pipe operator
i_7_sh <- 
  dat %>% 
  incidence(date_index = date_of_onset, interval = "week", 
            groups = c(gender, hospital))
i_7_sh
i_7_sh %>% summary()
i_7_sh %>% facet_plot(facets = gender, fill = hospital)
```


## Handling `incident` objects

`incidence` objects can be manipulated easily using both operators from base R
and dplyr.. The `[` operator implements normal subsetting of rows and columns.
For instance, to keep only the first 20 weeks of the epidemic:

```{r start}
plot(i_7[1:20, ])
```

incidence2 extends common dplyr verbs and this allows easy temporal subsetting:

```{r tail}
i_7 %>% 
  filter(date_group >= as.Date("2015-01-01")) %>% 
  plot(color = "white")
```

Subsetting groups can also matter. For instance, let's try and visualise the
incidence based on onset of symptoms by outcome:

```{r i7outcome}
i_7_outcome <- incidence(dat,
                         date_index = date_of_onset,
                         interval = 7, 
                         groups = outcome)
i_7_outcome
summary(i_7_outcome)
plot(i_7_outcome, stack = TRUE)
facet_plot(i_7_outcome, n_breaks = 3)
```

```{r i7outcome_cum}
i_7_outcome_cum <- cumulate(i_7_outcome)
i_7_outcome_cum
plot(i_7_outcome_cum)
```

Groups can also be collapsed into a single time series using `regroup`:

```{r regroup}
i_regrouped <- regroup(i_7_outcome)
i_regrouped
identical(i_7, i_regrouped)
```
