---
title: "Alternative date groupings"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Alternative date groupings}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 5
)
```

The `incidence()` function wraps the date grouping functionality of the [grates](https://cran.r-project.org/package=grates) package, providing an easy to use interface for constructing incidence objects.
Sometimes, however, you may want greater flexibility in choosing how you would like to transform your "date" inputs.
For this reason we also the provide the function `build_incidence()`, where you can specify the function you wish to apply.
We illustrate this approach using both the [clock](https://cran.r-project.org/package=clock) and [zoo](https://cran.r-project.org/package=zoo) packages.

```{r}
library(incidence2)
library(clock)
library(zoo)

# data included in incidence2 but obtained via the covidregionaldata package
data("covidregionaldataUK")
dat <- covidregionaldataUK
```

## Month grouping based on [clock](https://cran.r-project.org/package=clock) functions

```{r}
clock_month_inci <- 
  build_incidence(
    dat,
    date_index = date,
    groups = region,
    counts = ends_with("new"),
    FUN = function(x) calendar_narrow(as_year_month_day(x), precision = "month")
  )

clock_month_inci
```

## Month grouping [zoo](https://cran.r-project.org/package=zoo) function `as.yearmon`

```{r}
zoo_month_inci <- 
  build_incidence(
    dat,
    date_index = date,
    groups = region,
    counts = ends_with("new"),
    FUN = as.yearmon
  )

zoo_month_inci

```

## comparing the above to the [grates](https://cran.r-project.org/package=grates) implementation

```{r}
# we can compare this to the grates implementation
grates_month_inci <- 
  incidence(
    dat,
    date_index = date,
    groups = region,
    counts = ends_with("new"),
    interval = "month"
  )

grates_month_inci

# and check they are all "equal" ...
clock_month_inci$date_index <- as.Date(calendar_widen(clock_month_inci$date_index, "day"))
zoo_month_inci$date_index <- as.Date(zoo_month_inci$date_index)
grates_month_inci$date_index <- as.Date(grates_month_inci$date_index)

identical(
  as.data.frame(clock_month_inci),
  as.data.frame(zoo_month_inci)
)

identical(
  as.data.frame(grates_month_inci),
  as.data.frame(clock_month_inci)
)

```
