---
title: "Customizing incidence plots"
author: "Tim Taylor, Thibaut Jombart and Zhina N. Kamvar"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Customizing incidence plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center"
)
```

This vignette provides some tips for the most common customisations of graphics
produced by `plot.incidence()`. Our graphics use *ggplot2*, which is a distinct 
graphical system from base graphics. If you want advanced customisation of your
incidence plots, we recommend following an introduction to *ggplot2* such as
[R for Data Science](https://r4ds.had.co.nz/).

<br>


# Example data: simulated Ebola outbreak

This example uses the simulated Ebola Virus Disease (EVD) outbreak from the
package [*outbreaks*](http://github.com/reconhub/outbreaks): `ebola_sim_clean`.

First, we load the data, compute the weekly incidence and initially group by
gender and hospital:

```{r, data}
library(outbreaks)
library(incidence2)

dat <- ebola_sim_clean$linelist
str(dat)

i <- incidence(dat, date_index = date_of_onset, interval = 7,
               groups = c(gender, hospital))
```


# The `plot.incidence()` function

When calling `plot` on an `incidence()` object, the function `plot.incidence()`
is implicitly used. To access its documentation, use `?plot.incidence`. In this
section, we illustrate existing customisations.

## Default behaviour

By default, the function uses colors from the color palette `vibrant()`.  If no
fill is specified, groups will all be filled with the same color and a message
will be displayed due to multiple groups being present:

```{r, plot1}
i %>% plot()
```

Note, however, that the groups are still present just hidden by the default
border coloring:
```{r, plot2}
i %>% plot(color = "white")
```


