---
title: "Customizing incidence plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Customizing incidence plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 5
)
```

This vignette provides some tips for the most common customisations of graphics produced by `plot.incidence2()` and `facet_plot.incidence2()`.
Our graphics use [ggplot2](https://cran.r-project.org/package=ggplot2), which is a distinct graphical system from base graphics.

We try to provide reasonably good, out of the box, customisation via function arguments.
For advanced customisation, however, it may be better to first convert to a data.frame and then proceeding to work directly with [ggplot2](https://cran.r-project.org/package=ggplot2).

Note that plotting functionality only works with objects created with `incidence()` not `build_incidence()`.
This is due to `incidence()` using the [grates](https://cran.r-project.org/package=ggplot2) allowing us to rely on some of it's behaviour within out plotting functions.
For objects created with `build_incidence()` you can again use [ggplot2](https://cran.r-project.org/package=ggplot2) directly.

## Example data: simulated Ebola outbreak

This example uses the simulated Ebola Virus Disease (EVD) outbreak from the package [outbreaks](https://github.com/reconhub/outbreaks).

First, we load the data, compute the weekly incidence and initially group by gender and hospital:

```{r data, message = FALSE}
library(outbreaks)
library(incidence2)

dat <- ebola_sim_clean$linelist
str(dat)

i <- incidence(dat, date_of_onset, interval = 7, groups = c(gender, hospital))
i
```

## `plot.incidence2()`

When calling `plot` on an `incidence()` object, the function `plot.incidence2()` is implicitly used.
To access its documentation, use `?plot.incidence2`.
In this section, we illustrate existing customisations.

### Default behaviour

By default, the function uses colours from the colour palette `vibrant()`.
If no fill is specified, groups will all be filled with the same colour and a message will be displayed due to multiple groups being present:

```{r, plot1}
plot(i)
```

Note, however, that the groups are still present just hidden by the default border coloring:
```{r, plot2}
plot(i, color = "white")
```


### Additional arguments
`plot.incidence2()` is designed for a high level of customization without needing knowledge of [ggplot2](https://cran.r-project.org/package=ggplot2).
To this end, there are multiple arguments that can be provided which have been set to sensible defaults.
Some common changes you may wish to make are illustrated below (for a full description of arguments check out the accompanying help file `?plot.incidence2`).

We can easily fill the plot according to the groups present or a colour of our choosing:

```{r, fill}
plot(i, fill = gender)
plot(i, fill = hospital, legend = "bottom")
ii <- regroup(i)
plot(ii, fill = "red", color = "white")
```

Sometimes we may wish to change rotation of the x-axis and this can be done by passing an additional angle argument:

```{r, rotateandformat}
plot(i, angle = 45)
```

### Applying the style of European Programme for Intervention Epidemiology Training (EPIET)

For small datasets it is convention of EPIET to display individual cases as
rectangles. It can be done by doing two things: first, adding using the option
`show_cases = TRUE` with a white border. We also add `coord_equal = TRUE` which 
forces each case to be a square.

```{r, epiet}
i_epiet <- incidence(dat[160:180, ], date_index = date_of_onset)
plot(i_epiet, color = "white", show_cases = TRUE, angle = 45, size = 10, n.breaks = 20)
```


## `facet_plot()`

With stacked plots it can be difficult to ascertain difference between groups.
For this reason we provide users the ability to make faceted plots with the `facet_plot.incidence2()` function.
This function takes similar arguments to `plot` but includes an additional facet argument:

```{r, facets}
facet_plot(i, facets = gender, n.breaks = 3)
facet_plot(i, facets = hospital, fill = gender, n.breaks = 3, nrow = 4)
ii <- regroup(i, gender)
facet_plot(ii, facets = gender, fill = "grey", color = "white")
```

## Miscellaneous

### Changing colors

#### The default palette

A color palette is a function which outputs a specified number of colours.
By default, the colour used in *incidence* is called `vibrant()`.
Its behaviour is different from usual palettes, in the sense that the first 6 colours are not interpolated:

```{r, vibrant,   fig.height = 8}
par(mfrow = c(2, 1), mar = c(4,2,1,1))
barplot(1:6, col = vibrant(6))
barplot(1:20, col = vibrant(20))
```

We also provide a second palette called `muted()` where the first 9 colours are not interpolated:

```{r, muted, fig.height = 8}
par(mfrow = c(2,1), mar = c(4,2,1,1))
barplot(1:9, col = muted(9))
barplot(1:20, col = muted(20))
```

#### Using different palettes

Other color palettes can be provided via `col_pal`.
Various palettes are part of the base R distribution, and many more are provided in additional packages.
We provide a couple of examples below:

```{r, palettes}
ih <- regroup(i, hospital)
plot(ih, fill = hospital, col_pal = rainbow, n.breaks = 3) # see ?rainbow
ig <- regroup(i, gender)
plot(ig, fill = gender, col_pal = cm.colors)   # see ?cm.colors
```
