---
title: "Customizing incidence plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Customizing incidence plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 5
)
```

This vignette provides some tips for the most common customisations of graphics
produced by `plot.incidence()`. Our graphics use *ggplot2*, which is a distinct 
graphical system from base graphics. If you want advanced customisation of your
incidence plots, we recommend following an introduction to *ggplot2* such as
[R for Data Science](https://r4ds.had.co.nz/).

<br>


# Example data: simulated Ebola outbreak

This example uses the simulated Ebola Virus Disease (EVD) outbreak from the
package [*outbreaks*](https://github.com/reconhub/outbreaks): `ebola_sim_clean`.

First, we load the data, compute the weekly incidence and initially group by
gender and hospital:

```{r, data}
library(outbreaks)
library(incidence2)
library(magrittr)

dat <- ebola_sim_clean$linelist
str(dat)

i <- incidence(dat, date_index = date_of_onset, interval = 7,
               groups = c(gender, hospital))
i
```


# `plot.incidence()`

When calling `plot` on an `incidence()` object, the function `plot.incidence()`
is implicitly used. To access its documentation, use `?plot.incidence`. In this
section, we illustrate existing customisations.

## Default behaviour

By default, the function uses colors from the color palette `vibrant()`.  If no
fill is specified, groups will all be filled with the same color and a message
will be displayed due to multiple groups being present:

```{r, plot1}
i %>% plot()
```

Note, however, that the groups are still present just hidden by the default
border coloring:
```{r, plot2}
i %>% plot(color = "white")
```


## Additional arguments
`plot.incidence()` is designed for a high level of customization without needing
knowledge of *ggplot2*.  To this end, there are multiple arguments that can
be provided which have been set to sensible defaults. Some common changes you
may wish to make are illustrated below (for a full description of arguments
check out the accompanying help file `?plot.incidence`).

We can easily fill the plot according to the groups present or a color of our
choosing:

```{r, fill}
i %>% plot(fill = gender)
i %>% plot(fill = hospital, legend = "bottom")
i %>% regroup() %>% plot(fill = "red", color = "white")
```

Sometimes we may wish to change rotation of the x-axis and this can be done by
passing an additional angle argument:

```{r, rotateandformat}
i %>% plot(angle = 45)
```

## Applying the style of European Programme for Intervention Epidemiology Training (EPIET)

For small datasets it is convention of EPIET to display individual cases as
rectangles. It can be done by doing two things: first, adding using the option
`show_cases = TRUE` with a white border. We also add `coord_equal = TRUE` which 
forces each case to be a square.

```{r, epiet}
i_epiet <- incidence(dat[160:180, ], date_index = date_of_onset)
i_epiet %>% plot(color = "white", show_cases = TRUE, angle = 45, size = 10, n_breaks = 20)
```


# `facet_plot()`

With stacked plots it can be difficult to ascertain difference between groups.
For this reason we provide users the ability to make faceted plots with the
`facet_plot()` function.  This function takes similar arguments to `plot()` but
includes an additional facet argument:

```{r, facets}
i %>% facet_plot(facets = gender, n_breaks = 3)
i %>% facet_plot(facets = hospital, fill = gender, n_breaks = 3, nrow = 4)
i %>% regroup(gender) %>% facet_plot(facets = gender, fill = "grey", color = "white")
```


# Miscellaneous

## Changing colors

### The default palette

A color palette is a function which outputs a specified number of colors. By
default, the color used in *incidence* is called `vibrant()`. Its
behaviour is different from usual palettes, in the sense that the first 6
colours are not interpolated:

```{r, vibrant,   fig.height = 8}
par(mfrow = c(2, 1), mar = c(4,2,1,1))
barplot(1:6, col = vibrant(6))
barplot(1:20, col = vibrant(20))
```

We also provide a second palette called `muted()` where the first 9 colours are
not interpolated:

```{r, muted, fig.height = 8}
par(mfrow = c(2,1), mar = c(4,2,1,1))
barplot(1:9, col = muted(9))
barplot(1:20, col = muted(20))
```

### Using different palettes

Other color palettes can be provided via `col_pal`. Various palettes are part of
the base R distribution, and many more are provided in additional packages. We 
provide a couple of examples:

```{r, palettes}
i %>% regroup(hospital) %>% plot(fill = hospital, col_pal = rainbow, n_breaks = 3) # see ?rainbow
i %>% regroup(gender) %>% plot(fill = gender, col_pal = cm.colors)   # see ?cm.colors
```
