---
title: "Handling incidence objects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Handling incidence objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 5
)
```

`incidence()` objects are easy to work with, and we providing helper functions for both manipulating and accessing the underlying data and attributes.
As `incidence()` objects are subclasses of [tibbles](https://tibble.tidyverse.org/) they also have good integration with [tidyverse](https://www.tidyverse.org/) verbs.

## Modifying incidence objects

### `regroup()`
Sometimes you may find you've created a grouped `incidence()` but now want to change the internal grouping.
Assuming you are after a subset of the grouping already generated, then you can use to `regroup()` function to get the desired aggregation:

```{r regroup, message=FALSE}
library(outbreaks)
library(dplyr)
library(incidence2)

# load data
dat <- ebola_sim_clean$linelist

# generate the incidence object with 3 groups
inci <- incidence(dat, date_of_onset, groups = c(gender, hospital, outcome), interval = "week")
inci

# regroup to just two groups
inci %>% regroup(c(gender, outcome))

# drop all groups
inci %>% regroup()
```

### `keep_first()` and `keep_last()`
Once your data is grouped by date, you may want to select the first or last few
entries based on a particular date grouping using `keep_first()` and
`keep_last()`:

```{r keep}
inci %>% keep_first(3)
inci %>% keep_last(3)
```

### Tidyverse compatibility
incidence2 has been written with *tidyverse* compatibility (in particular [dplyr](https://dplyr.tidyverse.org/)) at the forefront of the design choices we have made.
By this we mean that if an operation from dplyr is applied to an `incidence()` object then as long as the invariants of the object are preserved (i.e. groups, interval and uniqueness of rows) then the object returned will be an `incidence()` object.
If the invariants are not preserved then a tibble will be returned instead.

```{r tidyverse}
library(dplyr)

# create incidence object
inci <- incidence(dat, date_of_onset, interval = "week", groups = c(hospital, gender))

# filtering preserves class
inci %>%  filter(gender == "f", hospital == "Rokupa Hospital")

# slice operations preserve class
inci %>% slice_sample(n = 10)

inci %>%  slice(1, 5, 10)

# mutate preserve class
inci %>%  mutate(future = date_index + 999)

# rename preserve class
inci %>%  rename(left_bin = date_index)

# select returns a tibble unless all date, count and group variables are preserved
inci %>% select(-1)

inci %>% select(everything())
```

# Accessing variable information
We provide multiple accessors to easily access information about an `incidence()` objects structure:

* `get_count_names()`, `get_dates_name()`, and `get_group_names()` all return character vectors of the column names corresponding to the requested variables.
* `get_n()` returns the number of observations.
* `get_interval()` returns the interval of the object.
* `get_timespan()` returns the number of days the object covers.
