---
title: "Handling incidence objects"
author: "Tim Taylor"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Handling incidence objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 5
)
```

We try to make `incidence()` objects easy to work with providing helper
functions for both manipulating and accessing data within the objects, 
and integration with [tidyverse](https://www.tidyverse.org/) verbs.

# Modifying incidence objects

## regroup
Sometimes you may find you've created a grouped `incidence()` but now want to
change the internal grouping.  Assuming you are after a subset of the grouping
already generated, then you can use to `regroup()` function to get the desired
aggregation:

```{r regroup}
library(outbreaks)
library(incidence2)

# load data
dat <- ebola_sim_clean$linelist

# generate the incidence object with 3 groups
inci <- incidence(dat, date_index = date_of_onset,
                  groups = c(gender, hospital, outcome),
                  interval = "1 week")
inci

# regroup to just two groups
inci %>% regroup(c(gender, outcome))

# drop all groups
inci %>% regroup()
```

## cumulate
We also provide a helper function, `cumulate`() to easily generate cumulative 
incidences:

```{r cumulate}
inci %>% 
  cumulate() %>% 
  facet_plot(facet = hospital, n_breaks = 2)
```


## Tidyverse compatibility
incidence2 has been written with *tidyverse* compatibility (in particular [dplyr](https://dplyr.tidyverse.org/)) at the forefront of the
design choices we have made.  By this we mean that if an operation from
dplyr is applied to an `incidence()` object then as long as the invariants of
the object are preserved (i.e. groups, interval and uniqueness of rows) then
the object returned will be an `incidence()` object.  If the invariants are
not preserved then a tibble will be returned instead.  Some examples of these
behaviours are given below:

```{r tidyverse}
library(dplyr)

# create incidence object
inci <-
  dat %>%
  incidence(
    date_index = date_of_onset,
    interval = "2 weeks",
    first_date = "2014-05-20",
    last_date = "2014-06-10",
    groups = c(hospital, gender)
  )

# filtering preserves class
x <-
    inci %>%
    filter(gender == "f", hospital == "Rokupa Hospital")
x
identical(class(x), class(inci))

# slice operations preserve class
x <-
    inci %>%
    slice_sample(n = 10)
x
identical(class(x), class(inci))

inci %>%
  slice(1, 5, 10) %>%
  class() %>% 
  identical(class(inci))

# mutate preserve class
x <-
    inci %>%
    mutate(future = bin_date + 999)
x
identical(class(x), class(inci))

# rename preserve class
x <-
  inci %>%
  rename(left_bin = bin_date)
identical(class(x), class(inci))

# select returns a tibble unless all date, count and group variables are preserved
inci %>% select(-1)

x <-
    inci %>%
    select(everything())
x

# Adding rows that are multiples of 2 weeks will maintain class
x <-
    inci %>%
    slice_head(n = 2) %>%
    mutate(bin_date = bin_date + 112) %>%
    bind_rows(inci)
x
identical(class(x), class(inci))


# Adding rows with dates that are not multiples of 2 weeks drops class
inci %>%
  slice_head(n = 2) %>%
  mutate(bin_date = bin_date + 30) %>%
  bind_rows(inci)
```

# Accessing variable information
We provide multiple accessors to easily access information about an `incidence()`
objects structure:

* `get_counts_name()`, `get_dates_name()`, `get_date_group_names()` and 
  `get_group_names()` all return character vectors of the column names corresponding
  to the requested variables.

* `get_n()` returns the number of observations.

* `get_interval()` returns the interval of the object.

* `get_timespan()`returns the number of days the object covers.




