---
title: "Handling incidence objects"
author: "Tim Taylor"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Handling incidence objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 5
)
```

We try to make `incidence()` objects easy to work with providing helper
functions for both manipulating and accessing data within the objects, 
and integration with [tidyverse](https://www.tidyverse.org/) verbs.

# Modifying incidence objects

## regroup
Sometimes you may find you've created a grouped `incidence()` but now want to
change the internal grouping.  Assuming you are after a subset of the grouping
already generated, then you can use to `regroup()` function to get the desired
aggregation:

```{r regroup}
library(outbreaks)
library(dplyr)
library(incidence2)

# load data
dat <- ebola_sim_clean$linelist

# generate the incidence object with 3 groups
inci <- incidence(dat, date_index = date_of_onset,
                  groups = c(gender, hospital, outcome),
                  interval = "yrwk")
inci

# regroup to just two groups
inci %>% regroup(c(gender, outcome))

# drop all groups
inci %>% regroup()
```

## cumulate
We also provide a helper function, `cumulate`() to easily generate cumulative 
incidences:

```{r cumulate}
inci %>% 
  regroup(hospital) %>% 
  cumulate() %>% 
  facet_plot(n_breaks = 4)
```


## Tidyverse compatibility
incidence2 has been written with *tidyverse* compatibility (in particular 
[dplyr](https://dplyr.tidyverse.org/)) at the forefront of the
design choices we have made.  By this we mean that if an operation from
dplyr is applied to an `incidence()` object then as long as the invariants of
the object are preserved (i.e. groups, interval and uniqueness of rows) then
the object returned will be an `incidence()` object.  If the invariants are
not preserved then a tibble will be returned instead.  Some examples of these
behaviours are given below:

```{r tidyverse}
library(dplyr)

# create incidence object
inci <-
  dat %>%
  incidence(
    date_index = date_of_onset,
    interval = "2 weeks",
    groups = c(hospital, gender)
  )

# filtering preserves class
inci %>%  filter(gender == "f", hospital == "Rokupa Hospital")

# slice operations preserve class
inci %>% slice_sample(n = 10)

inci %>%  slice(1, 5, 10)

# mutate preserve class
inci %>%  mutate(future = date_index + 999)

# rename preserve class
inci %>%  rename(left_bin = date_index)


# select returns a tibble unless all date, count and group variables are preserved
inci %>% select(-1)

inci %>% select(everything())

```

# Accessing variable information
We provide multiple accessors to easily access information about an `incidence()`
objects structure:

* `get_counts_name()`, `get_dates_name()`, `get_date_group_names()` and 
  `get_group_names()` all return character vectors of the column names corresponding
  to the requested variables.

* `get_n()` returns the number of observations.

* `get_interval()` returns the interval of the object.

* `get_timespan()`returns the number of days the object covers.




