---
title: "incidence2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{incidence2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 5
)
```

## What does it do?

**incidence2** is an R package that implements functions and classes to compute,
handle and visualise *incidence* data. It aims to be intuitive to use for both
interactive data exploration and as part of more robust outbreak analytic 
pipelines.

### Overview

The main feature of the package is the `incidence()` function which works on
both linelist and pre-aggregated data sets. It returns an object of class
`<incidence2>`. This is a data frame subclass with some additional invariants.
That is, an `<incidence2>` object must:
  
- have one column representing the date index (this does not need to be a `date`
  object but must have an inherent ordering over time);
  
- have one column representing the count variable (i.e. what is being counted)
  and one variable representing the associated count;
  
- have zero or more columns representing groups;

- not have duplicated rows with regards to the date and group variables.
    
To work with `<incidence2>` objects we provide a variety of functions:

- `plot.incidence2()`: generate simple plots with reasonable defaults.

- `regroup()`: regroup incidence from different groups into one global incidence
  time series.
  
- `keep_first()`, `keep_last()`: keep the rows corresponding to the first (or
  last) set of grouped dates (ordered by time) from an `<incidence>` object.
  
- `keep_peaks()`: keep the rows corresponding to the maximum count value for
  each grouping of an `<incidence>` object.

- `complete_dates()`: ensure every possible combination of date and groupings
  is represented with a count.
  
- `cumulate()`: calculate the cumulative incidence over time.
  
- `print.incidence2()` and `summary.incidence2()` methods.

- `as.data.frame.incidence2()` conversion method.

- Accessor functions for accessing underlying variables: `get_date_index()`,
  `get_count_variable()`, `get_count_value()`, `get_groups()`,
  `get_count_value_name()`, `get_count_variable_name()`, `get_date_index_name()`
  and `get_group_names()`.

### Computing and plotting incidence

```{r}
library(outbreaks)  # for the underlying data
library(ggplot2)    # For custom plotting later
library(incidence2) 
```

This example uses the simulated Ebola Virus Disease (EVD) outbreak from the
package [outbreaks](https://github.com/reconverse/outbreaks). 

```{r data}
dat <- ebola_sim_clean$linelist
str(dat)
```

To compute daily incidence we pass to `incidence()` a *data.frame* of
observation data. We must also pass the name of a variable in the data that we
can use to index the input. Note that whilst we we refer to this index as the
`date_index` there is no restriction on it's type save that it can be used to
represent the relative time of an observation:

```{r daily, fig.height = 5, dpi = 90}
daily <- incidence(dat, date_index = "date_of_onset")
daily
plot(daily)
```

The daily data is quite noisy, so we may want to pre group dates prior to
calculating the incidence. One way to do this is to utilise functions from the
[grates](https://www.reconverse.org/grates/dev/) package. Here we use the
`as_isoweek()` function to calculate the weekly (starting on a Monday)
incidence:

```{r}
# isoweek incidence
weekly_dat <- transform(dat, date_of_onset = as_isoweek(date_of_onset))
inci <- incidence(weekly_dat, date_index = "date_of_onset")
inci
plot(inci, border_colour = "white")
```

By grouping dates prior to calling `incidence()` it makes it clear to future
readers of your code (including yourself) which transformations are being
applied to your input data. However, as it is such a common and useful operation,
we have integrated much of [grates'](https://www.reconverse.org/grates/dev/)
functionality directly in to incidence2. This integration is done via an
`interval` parameter which can take values:

- week(s), isoweek(s) or weekly (mapping to `<grates_isoweek>`);
- epiweek(s) (mapping to `<grates_epiweek>`);
- month(s), yearmonth(s) or monthly (`<grates_yearmonth>`);
- quarter(s), yearquarter(s) or quarterly (`<grates_yearquarter>`);
- year(s) or yearly (`<grates_year>`).

As an example, the following is equivalent to the `inci` output above:

```{r}
# isoweek incidence using the interval parameter
inci2 <- incidence(dat, date_index = "date_of_onset", interval = "isoweek")
inci2

# check equivalent
identical(inci, inci2)
```

If we wish to aggregate by specified groups we can use the `groups` argument.
For instance, computing incidence by gender:

```{r}
inci_by_gender <- incidence(
    dat,
    date_index = "date_of_onset",
    groups = "gender",
    interval = "isoweek"
)
inci_by_gender
```

For grouped data, the plot method will create a faceted plot across groups:

```{r}
plot(inci_by_gender, border_colour = "white", angle = 45)
```

`incidence()` also supports multiple date inputs:
```{r}
grouped_inci <- incidence(
    dat,
    date_index = c(
        onset = "date_of_onset",
        infection = "date_of_infection"
    ), 
    interval = "isoweek",
    groups = "gender"
)
grouped_inci
```

When multiple date indices are given they are used for filling the resultant
plot:
```{r}
plot(grouped_inci, angle = 45, border_colour = "white")
```

Note that although the plot method for `<incidence2>` objects tries to balance
flexibility with simplicity it can often be preferable to call ggplot2 directly
on the underlying data.

### Modifying incidence objects

#### `regroup()`
Sometimes you may find you've created a grouped incidence but now want to change
the internal grouping. Assuming you are after a subset of the grouping already
generated, you can use `regroup()` to get the desired aggregation:

```{r}
# generate an incidence object with 3 groups
x <- incidence(
    dat,
    date_index = "date_of_onset",
    interval = "isoweek",
    groups = c("gender", "hospital", "outcome")
)

# regroup to just one group
xx <- regroup(x, c("gender", "outcome"))
xx

# drop all groups
regroup(x)
```

#### `cumulate()`
We also provide a helper function, `cumulate`() to easily generate cumulative 
incidences:

```{r}
y <- regroup(x, "hospital")
y <- cumulate(y)
y
plot(y, angle = 45, nrow = 3)
```

#### `keep_first()`, `keep_last()` and `keep_peaks()`
Once your data is grouped by date, you may want to select the first or last few
entries based on a particular date grouping using `keep_first()` and
`keep_last()`:

```{r}
keep_first(xx, 3)
keep_last(xx, 3)
```

Similarly you may want to quickly view the incidence peaks:

```{r}
keep_peaks(xx)
```


### subsetting and other manipulations
`<incidence>` objects have been carefully constructed to preserve their
structure under a range of different operations that can be applied to data
frames. By this we mean that if an operation is applied to an `<incidence>`
object then as long as the invariants of the object are preserved (i.e. groups,
interval and uniqueness of rows) then the object retain it's incidence class.
If the invariants are not preserved then a `<data frame>` will be returned
instead.

```{r subsetting}
inci <- incidence(
    dat,
    date_index = "date_of_onset",
    interval = "isoweek",
    groups = c("hospital", "gender")
)

# filtering preserves class
subset(inci, gender == "f" & hospital == "Rokupa Hospital")
inci[c(1L, 3L, 5L), ]

# Adding columns preserve class
inci$future <- inci$date_index + 999L
inci

# rename preserve class
names(inci)[names(inci) == "date_index"] <- "isoweek"
inci

# select returns a data frame unless all date, count and group variables are
# preserved in the output
str(inci[,-1L])
inci[, -6L]
```

### Accessing variable information

We provide multiple accessors to easily access information about an
`<incidence2>` object's structure:

```{r}
# the name of the date_index variable of x
get_date_index_name(inci)

# alias for `get_date_index_name()`
get_dates_name(inci)

# the name of the count variable of x
get_count_variable_name(inci)

# the name of the count value of x
get_count_value_name(inci)

# the name(s) of the group variable(s) of x
get_group_names(inci)

# list containing date_index variable of x
str(get_date_index(inci))

# alias for get_date_index
str(get_dates(inci))

# list containing the count variable of x
str(get_count_variable(inci))

# list containing count value of x
str(get_count_value(inci))

# list of the group variable(s) of x
str(get_groups(inci)) 
```

