---
title: "incidence2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{incidence2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 5
)
```

## What does it do?

{incidence2} is an R package that implements functions to compute, handle
and visualise *incidence* data. It aims to be intuitive to use for both
interactive data exploration and as part of more robust outbreak analytic
pipelines.

The package is based around objects of the namesake class, `<incidence2>`. These
objects are a data frame subclass with some additional invariants. That is, an
`<incidence2>` object must:
  
- have one column representing the date index (this does not need to be a `date`
  object but must have an inherent ordering over time);
  
- have one column representing the count variable (i.e. what is being counted)
  and one variable representing the associated count;
  
- have zero or more columns representing groups;

- not have duplicated rows with regards to the date and group variables.


To create and work with `<incidence2>` objects we provide a number of functions:

- `incidence()`: for the construction of incidence objects from both linelists
  and pre-aggregated data sets.

- `plot.incidence2()`: generate simple plots with reasonable defaults.

- `regroup()`: regroup incidence from different groups into one global incidence
  time series.
  
- `keep_first()`, `keep_last()`: keep the rows corresponding to the first (or
  last) set of grouped dates (ordered by time) from an `<incidence>` object.
  
- `keep_peaks()`: keep the rows corresponding to the maximum count value for
  each grouping of an `<incidence>` object.

- `complete_dates()`: ensure every possible combination of date and groupings
  is represented with a count.
  
- `cumulate()`: calculate the cumulative incidence over time.
  
- `print.incidence2()` and `summary.incidence2()` methods.

- `as.data.frame.incidence2()` conversion method.

- Accessor functions for accessing underlying variables: `get_date_index()`,
  `get_count_variable()`, `get_count_value()`, `get_groups()`,
  `get_count_value_name()`, `get_count_variable_name()`, `get_date_index_name()`
  and `get_group_names()`.
  
## Usage

The following sections give an overview of the package utilising two different
data sets. The first of these datasets comes from the
[{outbreaks}](https://cran.r-project.org/package=outbreaks) package and is a
synthetic linelist generated from a simulated Ebola Virus Disease (EVD)
outbreak. The second data set is available within {incidence2} and represents a
pre-aggregated time-series of Covid cases, tests, hospitalisations, and deaths
for UK regions that was obtained using the
[{covidregionaldata}](https://CRAN.R-project.org/package=covidregionaldata)
package (extracted on 2021-06-03).

```{r}
library(outbreaks)  # for the underlying data
library(ggplot2)    # For custom plotting later
library(incidence2) 

ebola <- ebola_sim_clean$linelist
str(ebola)
covid <- covidregionaldataUK
str(covid)
```

### Computing incidence from a linelist

To compute daily incidence we pass to `incidence()` a linelist of observation
data. This input should be in the form of a data frame and we must also pass the
name of a variable in the data that we can use to index the input. Note that
whilst we we refer to this index as the `date_index` there is no restriction on
it's type, save it needing represent the relative time of an observation
(i.e. it has an ordering).


```{r fig.height = 5, dpi = 90}
daily <- incidence(ebola, date_index = "date_of_onset")
daily
plot(daily)
```

The daily data is quite noisy, so we may want to pre group dates prior to
calculating the incidence. One way to do this is to utilise functions from the
[{grates}](https://www.reconverse.org/grates/dev/) package. Here we use the
`as_isoweek()` function to convert the 'date of onset' to an isoweek (a week
starting on a Monday) before calculating the incidence incidence:

```{r}
# isoweek incidence
weekly_ebola <- transform(ebola, date_of_onset = as_isoweek(date_of_onset))
inci <- incidence(weekly_ebola, date_index = "date_of_onset")
inci
plot(inci, border_colour = "white")
```

By grouping dates prior to calling `incidence()` it makes it clear to future
readers of your code (including yourself) which transformations are being
applied to your input data. This grouping, however, is such a common and useful
operation that we have chosen to integrate much of
[{grates}](https://www.reconverse.org/grates/dev/)
functionality directly in to incidence2. This integration is done via an
`interval` parameter in the `incidence()` call. This can take values:

- day or daily (mapping to `<Date>` objects);
- week(s), isoweek(s) or weekly (mapping to `<grates_isoweek>`);
- epiweek(s) (mapping to `<grates_epiweek>`);
- month(s), yearmonth(s) or monthly (`<grates_yearmonth>`);
- quarter(s), yearquarter(s) or quarterly (`<grates_yearquarter>`);
- year(s) or yearly (`<grates_year>`).

As an example, the following is equivalent to the `inci` output above:

```{r}
# isoweek incidence using the interval parameter
inci2 <- incidence(ebola, date_index = "date_of_onset", interval = "isoweek")
inci2

# check equivalent
identical(inci, inci2)
```

If we wish to aggregate by specified groups we can use the `groups` argument.
For instance, computing incidence by gender:

```{r}
inci_by_gender <- incidence(
    ebola,
    date_index = "date_of_onset",
    groups = "gender",
    interval = "isoweek"
)
inci_by_gender
```

For grouped data, the plot method will create a faceted plot across groups:

```{r}
plot(inci_by_gender, border_colour = "white", angle = 45)
```

`incidence()` also supports multiple date inputs:
```{r}
grouped_inci <- incidence(
    ebola,
    date_index = c(
        onset = "date_of_onset",
        infection = "date_of_infection"
    ), 
    interval = "isoweek",
    groups = "gender"
)
grouped_inci
```

When multiple date indices are given, they are used for filling the resultant
plot:
```{r}
plot(grouped_inci, angle = 45, border_colour = "white")
```

### Computing incidence from pre-aggregated data
The Covid data set is in a wide format with multiple count values given for each
day. To convert this to long form incidence we specify similar variables to
before but also include the count variables we are interested in:

```{r}
monthly_covid <- 
    covid |> 
    subset(!region %in% c("England", "Scotland", "Northern Ireland", "Wales")) |> 
    incidence(
        date_index = "date",
        groups = "region",
        counts = c("cases_new"),
        interval = "yearmonth"
    )
monthly_covid
plot(monthly_covid, nrow = 3, angle = 45, border_colour = "white")
```

### Modifying incidence objects

#### `regroup()`
Sometimes you may find you've created a grouped incidence but now want to change
the internal grouping. Assuming you are after a subset of the grouping already
generated, you can use `regroup()` to get the desired aggregation:

```{r}
# generate an incidence object with 3 groups
x <- incidence(
    ebola,
    date_index = "date_of_onset",
    interval = "isoweek",
    groups = c("gender", "hospital", "outcome")
)

# regroup to just one group
xx <- regroup(x, c("gender", "outcome"))
xx

# drop all groups
regroup(x)
```

#### `cumulate()`
We also provide a helper function, `cumulate`() to easily generate cumulative 
incidences:

```{r}
y <- regroup(x, "hospital")
y <- cumulate(y)
y
plot(y, angle = 45, nrow = 3)
```

### subsetting and other manipulations

#### `keep_first()`, `keep_last()` and `keep_peaks()`

Once your data is grouped by date, you may want to select the first or last few
entries based on a particular date grouping using `keep_first()` and
`keep_last()`:

```{r}
inci <- incidence(
    ebola,
    date_index = "date_of_onset",
    interval = "isoweek",
    groups = c("hospital", "gender")
)

keep_first(inci, 3)
keep_last(inci, 3)
```

Similarly you may want to quickly view the incidence peaks:

```{r}
keep_peaks(inci)
```

#### `complete_dates()`

Sometimes your incidence data does not span consecutive units of time, or
different groupings may cover different periods. To this end we provide a
`complete_dates()` function which ensures a complete and identical range of
dates are given counts (by default filling with a 0 value).

```{r}
dat <- data.frame(
    dates = as.Date(c("2020-01-01", "2020-01-04")),
    gender = c("male", "female")
)
i <- incidence(dat, date_index = "dates", groups = "gender")
i
complete_dates(i)
```

#### preservation of class

`<incidence2>` objects have been carefully constructed to preserve their
structure under a range of different operations that can be applied to data
frames. By this we mean that if an operation is applied to an `<incidence2>`
object then as long as the invariants of the object are preserved (i.e. groups,
interval and uniqueness of rows) then the object retain it's incidence class.
If the invariants are not preserved then a `<data.frame>` will be returned
instead.

```{r subsetting}

# filtering preserves class
subset(inci, gender == "f" & hospital == "Rokupa Hospital")
inci[c(1L, 3L, 5L), ]

# Adding columns preserve class
inci$future <- inci$date_index + 999L
inci

# rename preserve class
names(inci)[names(inci) == "date_index"] <- "isoweek"
inci

# select returns a data frame unless all date, count and group variables are
# preserved in the output
str(inci[,-1L])
inci[, -6L]
```

### Accessing variable information

We provide multiple accessors to easily access information about an
`<incidence2>` object's structure:

```{r}
# the name of the date_index variable of x
get_date_index_name(inci)

# alias for `get_date_index_name()`
get_dates_name(inci)

# the name of the count variable of x
get_count_variable_name(inci)

# the name of the count value of x
get_count_value_name(inci)

# the name(s) of the group variable(s) of x
get_group_names(inci)

# list containing date_index variable of x
str(get_date_index(inci))

# alias for get_date_index
str(get_dates(inci))

# list containing the count variable of x
str(get_count_variable(inci))

# list containing count value of x
str(get_count_value(inci))

# list of the group variable(s) of x
str(get_groups(inci)) 
```

