---
title: "An introduction to incidence2"
output:
    markdown::html_format:
vignette: >
  %\VignetteIndexEntry{An introduction to incidence2}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{dplyr, tidyr}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center"
)
data.table::setDTthreads(2)
```

## What does it do?

**incidence2** is an R package that implements functions to compute, handle
and visualise *incidence* data. It aims to be intuitive to use for both
interactive data exploration and as part of more robust outbreak analytic
pipelines.

The package is based around objects of the namesake class, `incidence2`. These
objects are a data frame subclass with some additional invariants. That is, an
`incidence2` object must:
  
- have one column representing the date index (this does not need to be a `Date`
  object but must have an inherent ordering over time);
  
- have one column representing the count variable (i.e. what is being counted)
  and one variable representing the associated count;
  
- have zero or more columns representing groups;

- not have duplicated rows with regards to the date, group and count variables.

## Usage

To construct `incidence2` objects we utilise the, imaginatively named,
`incidence()` function. This works with data frames that can be in either
*linelist* or *pre-aggregated* format. Broadly speaking, we refer to data with
one row of observations (e.g. "Sex", "Date of symptom onset",
"Date of Hospitalisation") per individual as a linelist. In terms of this
package, pre-aggregated data, is data where we have a single column representing
time and associated counts linked to those times (still optionally split by
characteristics).

The following sections give an overview of how the package handles the two
different input formats. Firstly using synthetic linelist data generated from a
simulated Ebola Virus Disease (EVD) outbreak and available in the 
[outbreaks](https://cran.r-project.org/package=outbreaks) package. Secondly,
a pre-aggregated dataset (contained within incidence2) that contains a
time-series of Covid cases, tests, hospitalisations, and deaths for UK regions.

```{r setup}
library(incidence2)
library(outbreaks)
```

### Computing incidence from a linelist

To compute daily incidence we pass to `incidence()` our linelist of observation
data as well as the name of a column in the data that we can use to index over
time. Note that whilst we we refer to this index as the `date_index` there is no
restriction on it's type, save it needing represent the relative time of an
observation (i.e. it has an ordering).

### Computing incidence from a linelist

To compute daily incidence we pass to `incidence()` a linelist of observation
data. This input should be in the form of a data frame and we must also pass the
name of a variable in the data that we can use to index the input. Note that
whilst we we refer to this index as the `date_index` there is no restriction on
it's type, save it needing represent the relative time of an observation
(i.e. it has an ordering).

First we get the linelist from the simulated outbreak data and calculate the
daily incidence.

```{r}
ebola <- dplyr::filter(ebola_sim_clean$linelist ,!is.na(hospital))
str(ebola)
(daily_incidence <- incidence(ebola, date_index = "date_of_onset"))
```

incidence2 also provides a simple plot function (see `help("plot.incidence2")`)
for visualising the resulting object. This functionality is built upon
[ggplot2](https://cran.r-project.org/package=ggplot2) and whilst it is useful
for quick interactive exploration of the data, due to the complexities of
automating plots it is often better to work with ggplot2 directly.

```{r fig.height = 5, dpi = 90}
plot(daily_incidence)
```

The daily data is quite noisy, so it may be worth grouping the dates prior to
calculating the incidence. One way to do this is to utilise functions from the
[grates](https://cran.r-project.org/package=grates) package. incidence2
depends on the grates package so all of it's functionality is available
directly to users. Here we use the `as_isoweek()` function to convert the
'date of onset' to an isoweek (a week starting on a Monday) before proceeding
to calculate the incidence:

```{r}
weekly_ebola <- dplyr::mutate(ebola, date_of_onset = as_isoweek(date_of_onset))
(weekly_incidence <- incidence(weekly_ebola, date_index = "date_of_onset"))
plot(weekly_incidence, border_colour = "white")
```

This sort of date grouping, is such a common and useful operation that we have
chosen to integrate this functionality directly in to `incidence()` via an
`interval` parameter. `interval` can take any of the following values:

- day or daily (mapping to `<Date>` objects);
- week(s), isoweek(s) or weekly (mapping to `<grates_isoweek>`);
- epiweek(s) (mapping to `<grates_epiweek>`);
- month(s), yearmonth(s) or monthly (`<grates_yearmonth>`);
- quarter(s), yearquarter(s) or quarterly (`<grates_yearquarter>`);
- year(s) or yearly (`<grates_year>`).

As an example, the following is equivalent to the `weekly_incidence` output above:

```{r}
(dat <- incidence(ebola, date_index = "date_of_onset", interval = "isoweek"))
# check equivalent
identical(dat, weekly_incidence)
```

If we wish to aggregate by specified groups we can use the `groups` argument.
For instance, to compute the weekly incidence by gender:

```{r}
(weekly_incidence_gender <- incidence(
    ebola,
    date_index = "date_of_onset",
    groups = "gender",
    interval = "isoweek"
))
```

For grouped data, the plot method will create a faceted plot across groups
unless a fill variable is specified:

```{r}
plot(weekly_incidence_gender, border_colour = "white", angle = 45)
plot(weekly_incidence_gender, border_colour = "white", angle = 45, fill = "gender")
```

`incidence()` also supports multiple date inputs:
```{r}
(weekly_multi_dates <- incidence(
    ebola,
    date_index = c(
        onset = "date_of_onset",
        infection = "date_of_infection"
    ), 
    interval = "isoweek",
    groups = "gender"
))
```

When multiple date indices are given, they are used for rows of the resultant
plot, unless the resultant variable is used to fill:
```{r}
plot(weekly_multi_dates, angle = 45, border_colour = "white")
plot(weekly_multi_dates, angle = 45, border_colour = "white", fill = "count_variable")
```

We also provide a `summary.incidence2()` method which can be useful for  a quick,
high-level, overview of the data

```{r}
summary(weekly_multi_dates)
```

#### Plotting in style of European Programme for Intervention Epidemiology Training (EPIET)

For small datasets it is convention of EPIET to display individual cases as
rectangles. We can do this by setting `show_cases = TRUE` in the call to
`plot()` which will display each case as an individual square with a white
border.

```{r, fig.height=3}
dat <- ebola[160:180, ]

incidence(
    dat,
    date_index = "date_of_onset",
    date_names_to = "date"
) |> 
plot(color = "white", show_cases = TRUE, angle = 45, n_breaks = 10)
incidence(
    dat,
    date_index = "date_of_onset",
    groups = "gender",
    date_names_to = "date"
) |> 
plot(show_cases = TRUE, color = "white", angle = 45, n_breaks = 10, fill = "gender")
```

### Computing incidence from pre-aggregated data

```{r}
covid <- dplyr::filter(
    covidregionaldataUK,
    !region %in% c("England", "Scotland", "Northern Ireland", "Wales")
)
str(covid)
```

The Covid data set is in a wide format with multiple count values given for each
day. Like with our linelist data, `incidence()` requires us to specify a
`date_index` column and optionally `groups` and `interval`. In addition we must
now also provide the `counts` variable(s) that we are interested in. Before
continuing, take note of the missing values, in the Covid data above. If these
occur in one of the count variables, `incidence()` will warn users:

```{r}
monthly_covid <- incidence(
    covid,
    date_index = "date",
    groups = "region",
    counts = "cases_new",
    interval = "yearmonth"
)

monthly_covid
```

Whilst we could have let `incidence()` ignore missing values (equivalent to
setting `sum(..., na.rm=TRUE)`), we prefer that users make an explicit choice
on how these should (or should not) be imputed. For example, to treat missing
values as zero counts we can simply replace them in the data prior to calling
`incidence()`:

```{r}
(monthly_covid <-
     covid |>
     tidyr::replace_na(list(cases_new = 0)) |> 
     incidence(
         date_index = "date",
         groups = "region",
         counts = "cases_new",
         interval = "yearmonth"
     ))
plot(monthly_covid, nrow = 3, angle = 45, border_colour = "white")
```


## Next steps

- `vignette(topic = "working-with-incidence2-objects", package = "incidence2")`
