---
title: "Introduction to the incidence2 package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to the incidence2 package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 5
)
```

## What does it do?
**incidence2** is an R package that implements functions and classes to compute, handle and visualise *incidence* from linelist data.

The main features of the package include:

* The `incidence()` and `build_incidence()` functions compute *incidence* from both linelist and pre-aggregated datasets across a range of date groupings.
  The returned object from `incidence()` is a subclass of [tibble](https://CRAN.R-project.org/package=tibble).
  This is compatible with [dplyr](https://dplyr.tidyverse.org/) for data manipulation (see `vignette("handling_incidence_objects")` for more details).
* Plotting methods, `plot.incidence2()` and `facet_plot.incidence2()` that provide quick plots with sensible defaults.
* Additional functions for handling incidence objects, including:
  - `regroup()`: regroup incidence from different groups into one global incidence time series.
  - `keep_first()` and `keep_last()` : will keep the rows corresponding to the first (or last) set of grouped dates (ordered by time) from an `incidence()` object.
  - `complete_counts()`: ensure every possible combination of date and groupings is given an explicit count.
  - `print.incidence_df()`** and `summary.incidence_df()` methods.
  - `as.data.frame.incidence_df()` and `as_tibble.incidence_df()` conversion methods.
  - Accessor functions: `get_count_names()`, `get_dates_name()`, `get_date_index()`, `get_group_names()`, `get_interval()`, `get_timespan()` and  `get_n()`.
  

## Worked example: simulated Ebola outbreak

This example uses the simulated Ebola Virus Disease (EVD) outbreak from the package [outbreaks](https://github.com/reconverse/outbreaks).
We will compute incidence for various time steps and illustrate how to easily plot the data.

### Loading the data

```{r data}
library(outbreaks)
library(incidence2)
dat <- ebola_sim_clean$linelist
class(dat)
str(dat)
```

### Computing and plotting incidence
To compute daily incidence we must pass observation data in the form of a *data.frame* to `incidence()`.
We must also pass the name of a date variable in the data that we can use to index the input:

First compute the daily incidence:
```{r daily, fig.height = 5, dpi = 90}
daily <- incidence(dat, date_index = date_of_onset)
daily
summary(daily)
plot(daily)
```

The daily incidence is quite noisy, but we can easily compute other incidence using other time intervals

```{r sevenday}
# 7 day incidence
seven <- incidence(dat, date_index = date_of_onset, interval = 7)
seven
plot(seven, color = "white")
```
Notice how specifying the interval as 7 creates weekly intervals with the coverage displayed by date.
Below we illustrate how `incidence()` also allows us to create year-weekly groupings with the default being weeks starting on a Monday (following the ISO 8601 date and time standard).

```{r weekly}
# year-weekly, starting on Monday (ISO week, default)
weekly <- incidence(dat, date_index = date_of_onset, interval = "week")
plot(weekly, color = "white")
```

`incidence()` will also work with larger intervals

```{r otherinterval}
# bi-weekly, based on first day in data
biweekly <- incidence(dat, date_index = date_of_onset, interval = "2 weeks")
plot(biweekly, color = "white")
# monthly
monthly <- incidence(dat, date_index = date_of_onset, interval = "month")
plot(monthly, color = "white")
# quarterly
quarterly <- incidence(dat, date_index = date_of_onset, interval = "quarter")
plot(quarterly, color = "white")
# year
yearly <- incidence(dat, date_index = date_of_onset, interval = "year")
plot(yearly, color = "white", n_breaks = 2)
```

### Grouping
`incidence()` can also aggregate incidence by specified groups using the `groups` argument.
For instance, we can compute incidence by gender and plot with both the `plot.incidence_df()` function for a single or the `facet_plot.incidence_df()` function for a multi-faceted plot across groups:

```{r gender}
weekly_grouped <- incidence(dat, date_of_onset, interval = "week", groups = gender)
weekly_grouped
summary(weekly_grouped)
# A singular plot
plot(weekly_grouped, fill = gender, color = "white")
# a multi-facet plot
facet_plot(weekly_grouped, fill = gender, n_breaks = 5, angle = 45, color = "white")
```

There is no limit to the number of groups that we group by and this allows us to both facet and fill by different variables:
```{r outcome_hospital}
inci <- incidence(dat, date_of_onset, interval = "week", groups = c(outcome, hospital))
facet_plot(inci, facets = hospital, fill = outcome, nrow = 3, n_breaks = 5, angle = 45)
```
