---
title: "Introduction to the incidence package"
author: "Tim Taylor, Thibaut Jombart and Zhina N. Kamvar"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to the incidence package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 5
)
```

*incidence* implements functions and classes to compute, handle and visualise 
incidences from linelist data. This vignette provides an introduction to current
features and is more extensive than the overview provided in README.md.


# Installing the package

You can install the current version of the package from 
[CRAN](https://cran.r-project.org/) with:

```{r, eval=FALSE}
install.packages("incidence2")
```

The development version can be installed from [GitHub](https://github.com/) with:

```{r, eval=FALSE}
if (!require(remotes)) {
  install.packages("remotes")
}
remotes::install_github("reconhub/incidence2", build_vignettes = TRUE)
```


# What does it do?

The main features of the package include:

- **`incidence()`**: compute incidence from both linelist and pre-aggregated
  datasets; any fixed time interval can be used; the returned object is a 
  tibble subclass called *incidence2*.

- plotting functions **`plot()`** and **`facet_plot()`**: these functions
  return customised ggplot2 plots of *incidence2* objects 
  (see **`plot.incidence2()`** for details).

- Compatible with [dplyr](https://dplyr.tidyverse.org/) for data manipulation.
  (see `vignette("handling_incidence_objects")` for more details).

- **`regroup()`**: regroup incidence from different groups into one global incidence
  time series.

- **`cumulate()`**: computes cumulative incidence over time from an `incidence()`
  object.

- **`print()`** and **`summary()`** functions.

- Conversion functions:
    
    - **`as.data.frame()`**: converts an `incidence()` object into a `data.frame`.
    
    - **`as_tibble()`**: converts an `incidence()` object into a `tibble`.

- Accessor functions: **`get_counts_name()`**, **`get_dates_name()`**,
  **`get_dates()`**, **`get_group_names()`**, **`get_interval()`**, 
  **`get_timespan()`** and  **`get_n()`**.


# Worked example: simulated Ebola outbreak

This example uses the simulated Ebola Virus Disease (EVD) outbreak from the package
[*outbreaks*](https://github.com/reconhub/outbreaks). We will compute incidence
for various time steps and illustrate how to easily plot the data.

## Loading the data

```{r data}
library(outbreaks)
library(incidence2)
library(magrittr)

dat <- ebola_sim_clean$linelist
class(dat)
str(dat)
```

## Computing and plotting incidence
To compute daily incidence we must pass observation data in the form of a
*data.frame* or *tibble* to the `incidence()`. We must also pass the name of
a date variable in the data that we can use to index the input:

We compute the daily incidence:
```{r incid1}
i <- incidence(dat, date_index = date_of_onset)
i
summary(i)
plot(i)
```

The daily incidence is quite noisy, but we can easily compute other incidence
using larger time intervals:

```{r interv}

# year-weekly, starting on Monday (ISO week, default)
i_weekly <- incidence(dat, date_index = date_of_onset, interval = "week")
plot(i_weekly)

# bi-weekly, based on first day in data
i_biweekly <- incidence(dat, date_index = date_of_onset, interval = "2 weeks")
plot(i_biweekly)

## year-monthly
i_monthly <- incidence(dat, date_index = date_of_onset, interval = "month")
plot(i_monthly)
```

## Grouping
`incidence()` can also aggregate incidence by specified groups using the
`groups` argument. For instance, we can compute incidence by gender and plot
with both the `plot()` function for a single plot and the `facet_plot()`
function for a multi-faceted plot across groups:

```{r gender}
i_weekly_sex <- incidence(dat, date_index = date_of_onset,
                     interval = "week", groups = gender)
i_weekly_sex
summary(i_weekly_sex)

# A singular plot
plot(i_weekly_sex, fill = gender)

# a multi-facet plot
facet_plot(i_weekly_sex, fill = gender, n_breaks = 5, angle = 45)
```

There is no limit to the number of groups that we group by and this allows
us to both facet and fill by different variables:
```{r outcome_hospital}
inci <- incidence(dat, date_index = date_of_onset,
                     interval = "week", groups = c(outcome, hospital, gender))
inci %>% 
  facet_plot(facets = hospital, fill = outcome, nrow = 3, n_breaks = 5, angle = 45)
```
