---
title: "Working with incidence2 objects"
output:
    markdown::html_format:
vignette: >
  %\VignetteIndexEntry{Working with incidence2 objects}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center"
)
data.table::setDTthreads(2)
```

On top of the incidence constructor function and the plotting / printing and
summary methods seen in the introductory vignette we provide a number of other
useful functions and integrations for working with incidence2 objects:

- `regroup()`: regroup incidence from different groups into one global incidence
  time series.
  
- `cumulate()`: calculate the cumulative incidence over time.
  
- `complete_dates()`: ensure every possible combination of date and groupings
  is represented with a count.

- `keep_first()`, `keep_last()`: keep the rows corresponding to the first (or
  last) set of grouped dates (ordered by time) from an `<incidence>` object.

- `keep_peaks()`; keep the rows corresponding to the maximum count value for
  each grouping of an `<incidence>` object. A convenience wrapper around this,
  `first_peak()` keeps returns the earliest occurring peak row.
  
- `bootstrap_incidence()`; sampling (with replacement and optional randomisation)
  from incidence2 objects.
  
- `estimate_peak()`; estimate the peak of an epidemic curve using bootstrapped
  samples of the available data.

- An `as.data.frame.incidence2()` conversion method.

- Accessors for accessing underlying variables: `get_date_index()`,
  `get_count_variable()`, `get_count_value()`, `get_groups()`,
  `get_count_value_name()`, `get_count_variable_name()`, `get_date_index_name()`
  and `get_group_names()`.
  
## Additional functions

Like in the introductory vignette we will use the synthetic linelist data
generated from a simulated Ebola Virus Disease (EVD) outbreak as well as some
data on 136 cases of influenza A H7N9 in China. Both data sets are available in
the [outbreaks](https://cran.r-project.org/package=outbreaks) package.

```{r setup}
library(outbreaks)
library(incidence2)
ebola <- dplyr::filter(ebola_sim_clean$linelist ,!is.na(hospital))
str(ebola)
```

### `regroup()`

If you've created a grouped incidence but now want to change the internal
grouping you can `regroup()` to get the desired aggregation:

```{r}
# generate an incidence object with 3 groups
(x <- incidence(
    ebola,
    date_index = "date_of_onset",
    interval = "isoweek",
    groups = c("gender", "hospital", "outcome")
))
# regroup to just two groups
regroup(x, c("gender", "outcome"))
# drop all groups
regroup(x)
```

### `cumulate()`
You can use `cumulate()` to easily generate cumulative incidences:

```{r}
(y <- cumulate(regroup(x, "hospital")))
plot(y, angle = 45, nrow = 3)
```

### `complete_dates()`
Sometimes your incidence data does not span consecutive units of time, or
different groupings may cover different periods. To this end we provide a
`complete_dates()` function which ensures a complete and identical range of
dates are given counts (by default filling with a 0 value).

```{r}
dat <- data.frame(
    dates = as.Date(c("2020-01-01", "2020-01-04")),
    gender = c("male", "female")
)

(incidence <- incidence(dat, date_index = "dates", groups = "gender"))
complete_dates(incidence)
```

### `keep_first()`, `keep_last()` and `keep_peaks()`

Once your data is grouped by date, you may want to select the first or last few
entries based on a particular date grouping using `keep_first()` and
`keep_last()`:

```{r}
weekly_incidence <- incidence(
    ebola,
    date_index = "date_of_onset",
    interval = "isoweek",
    groups = "hospital"
)

keep_first(weekly_incidence, 3)
keep_last(weekly_incidence, 3)
keep_peaks(weekly_incidence)
```

### Bootstrapping and estimating peaks

We provide functions to return the first peak of the incidence data, bootstrap
from the data and, wrapping these two functions, estimate confidence intervals
around a peak.

Note that the bootstrapping approach used for estimating the peak time makes
the following assumptions:

- the total number of event is known (no uncertainty on total incidence);
- dates with no events (zero incidence) will never be in bootstrapped datasets; and
- the reporting is assumed to be constant over time, i.e. every case is
  equally likely to be reported.


```{r}
dat <- fluH7N9_china_2013
influenza <- incidence(dat, date_index = "date_of_onset", groups = "province")

# across provinces (we suspend progress bar for markdown)
estimate_peak(influenza, progress = FALSE) |> 
    dplyr::select(-count_variable)
# regrouping for overall peak
plot(regroup(influenza))
estimate_peak(regroup(influenza), progress = FALSE) |> 
    dplyr::select(-count_variable)
# return the first peak of the grouped and ungrouped data
first_peak(influenza)
first_peak(regroup(influenza))
# bootstrap a single sample
bootstrap_incidence(influenza)
```


## Additional integrations

### Preservation of class

incidence2 objects have been carefully constructed to preserve their structure
under a range of different operations that can be applied to data frames. By
this we mean that if an operation is applied to an incidence2 object then as
long as the invariants of the object are preserved (i.e. groups, interval and
uniqueness of rows) then the object retain it's incidence class. If the
invariants are not preserved then a `data.frame` will be returned
instead.

```{r}

weekly_incidence <- incidence(
    ebola,
    date_index = "date_of_onset",
    interval = "isoweek",
    groups = c("gender", "hospital")
)

# filtering preserves class
subset(weekly_incidence, gender == "f" & hospital == "Rokupa Hospital")
weekly_incidence[c(1L, 3L, 5L), ]
# Adding columns preserve class
weekly_incidence$future <- weekly_incidence$date_index + 999L
weekly_incidence
# rename preserve class
names(weekly_incidence)[names(weekly_incidence) == "date_index"] <- "isoweek"
weekly_incidence
# select returns a data frame unless all date, count and group variables are
# preserved in the output
str(weekly_incidence[,-1L])
weekly_incidence[, -6L]
```

### Accessing variable information

We provide multiple accessors to easily access information about an
`<incidence2>` object's structure:

```{r}
# the name of the date_index variable of x
get_date_index_name(weekly_incidence)
# alias for `get_date_index_name()`
get_dates_name(weekly_incidence)
# the name of the count variable of x
get_count_variable_name(weekly_incidence)
# the name of the count value of x
get_count_value_name(weekly_incidence)
# the name(s) of the group variable(s) of x
get_group_names(weekly_incidence)
# the date_index variable of x
str(get_date_index(weekly_incidence))
# alias for get_date_index
str(get_dates(weekly_incidence))
# the count variable of x
str(get_count_variable(weekly_incidence))
# the count value of x
str(get_count_value(weekly_incidence))
# list of the group variable(s) of x
str(get_groups(weekly_incidence))
```
